Reducing Acceptance Marks in Emerson-Lei Automata by QBF-solving

### 2022-03-01 paper/main.tex
some figures and more text in sec. 3

In graphical representation of automata, we often use acceptance marks
$\tacc{1}{1},\tacc{2}{2},\ldots\in M$. An edge 
\begin{tikzpicture}[smallautomaton,baseline=-2pt]
  \node[state] (p) at (0,0) {$p$};
  \node[state] (q) at (1.8,0) {$q$};
  \path [->] (p) edge node[above,pos=.5,overlay] {$a$} node[accset,collacc8, pos=0.65] {k} node[accset,collacc4, pos=0.3] {j} (q);
\end{tikzpicture}
denotes the transition $(p,a,\{\tacc{4}{j},\tacc{8}{k}\},q)\in\delta$.

@@

In the following, we define three versions of $\mcycle(\vec{e})$:
\begin{itemize}
\item $\mcycle_1(\vec{e})$ is a lightweight version, which puts
  only minimal requirements on $\Te$. It does not use the
  information about automaton structure, but it comes with an
  interesting simplification of the whole formula $\Phi_{C,K}$.
\item $\mcycle_2(\vec{e})$ is a balanced version. It says that $\Te$
  is nonempty and every transition in $\Te$ has to have a preceding
  and a succeeding transition in $\Te$, which is a necessary condition
  for being a cycle, but not a sufficient one.
\item $\mcycle_3(\vec{e})$ is a strict version saying that $\Te$ is a
  cycle. Unfortunately, it uses additional universally quantified
  variables corresponding to automata states. Transformation to prenex
  normal form turns the quantifiers to existential ones and the
  resulting formula thus contains quantifier alternation.
\end{itemize}

@@

\subsection{Balanced version $\mcycle_2(\vec{e})$}

\subsection{Strict version $\mcycle_3(\vec{e})$}

>>>>

In graphical representation of automata, we often use acceptance marks
$\tacc{1}{1},\tacc{2}{2},\ldots\in M$. Further, an edge 
\begin{tikzpicture}[smallautomaton,baseline=-2pt]
  \node[state] (p) at (0,0) {$p$};
  \node[state] (q) at (1.8,0) {$q$};
  \path [->] (p) edge node[above,pos=.5,overlay] {$a$} node[accset,collacc8, pos=0.65] {k} node[accset,collacc4, pos=0.3] {j} (q);
\end{tikzpicture}
denotes the transition $(p,a,\{\tacc{4}{j},\tacc{8}{k}\},q)\in\delta$.

@@

In the following, we define three versions of $\mcycle(\vec{e})$:
\begin{itemize}
\item $\mcycle_1(\vec{e})$ is a lightweight version, which only says
  that $\Te$ is nonempty. It does not use the information about
  automaton structure, but it comes with an interesting simplification
  of the whole formula $\Phi_{C,K}$.
\item $\mcycle_2(\vec{e})$ is an intermediate version. It says that $\Te$
  is nonempty and every transition in $\Te$ has to have a preceding
  and a succeeding transition in $\Te$, which is a necessary condition
  for being a cycle, but not a sufficient one.
\item $\mcycle_3(\vec{e})$ is a strict version saying that $\Te$ is a
  cycle. Unfortunately, it uses additional universally quantified
  variables corresponding to automata states. Transformation of
  $\Phi_{C,K}$ to prenex normal form turns the quantifiers to
  existential ones and the resulting formula thus contains quantifier
  alternation.
\end{itemize}
We write $\Phi_{i,C,K}$ when we want to emphasize that the particular
formula $\Phi_{C,K}$ contains the version $\mcycle_i(\vec{e})$.

@@

\subsection{Intermediate version $\mcycle_2(\vec{e})$}
The intermediate version says that $\Te$ is nonempty and for each state
$q\in Q$ it holds that $\Te$ contains a transition leading to $q$ if
and only if it contains a transition leading from $q$.
Formally,
\[
  \mcycle_2(\vec{e})=\bigvee_{t\in\delta}e_t~\wedge~
  \bigwedge_{q\in Q}\Big(\bigvee_{t'\in\delta~\cap~Q\times\Sigma\times 2^M\times\{q\}}\hspace{-1cm}e_{t'}\hspace{1cm}\iff
  \bigvee_{t''\in\delta~\cap~\{q\}\times\Sigma\times 2^M\times Q}\hspace{-1cm}e_{t''}\hspace{6mm}\Big).
\]
This condition is satisfied by every cycle, but also by some sets of
transitions that are not cycles. Some examples of such sets are provided in Figure~\ref{fig:inter}.

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[smallautomaton]
    \begin{scope}[initial angle=90,xshift=-5cm]
      \node[state,initial] (p) at (0,0) {};
      \node[state] (q) at (1.8,0) {};
      \path[->,auto]
      (p) edge[out=-150,in=150,looseness=10] (p)
      (p) edge[out=30,in=150] (q)
      (q) edge[out=30,in=-30,looseness=10] (q)
      (q) edge[out=-150,in=-30] (p);
    \end{scope}

    \begin{scope}[initial angle=90,densely dotted,black!20]
      \node[state,initial,solid,draw=black] (p) at (0,0) {};
      \node[state,solid,draw=black] (q) at (1.8,0) {};
      \path[->,auto]
      (p) edge[out=-150,in=150,looseness=10,black,solid] (p)
      (p) edge[out=30,in=150,black,solid] (q)
      (q) edge[out=30,in=-30,looseness=10,black,solid] (q)
      (q) edge[out=-150,in=-30] (p);
    \end{scope}

    \begin{scope}[initial angle=90,densely dotted,black!20,xshift=5cm]
      \node[state,initial,solid,draw=black] (p) at (0,0) {};
      \node[state,solid,draw=black] (q) at (1.8,0) {};
      \path[->,auto]
      (p) edge[out=-150,in=150,looseness=10,black,solid] (p)
      (p) edge[out=30,in=150] (q)
      (q) edge[out=30,in=-30,looseness=10,black,solid] (q)
      (q) edge[out=-150,in=-30] (p);
    \end{scope}
  \end{tikzpicture}
  \caption{An automaton structure where transition labels and
    acceptance marks are not depicted (left) and two sets $\Te$ that
    are not cycles even if $\mcycle_2(\vec{e})$ holds (middle and
    right)}
  \label{fig:inter}
\end{figure}

\subsection{Strict version $\mcycle_3(\vec{e})$}
Before we give a definition of $\mcycle_3(\vec{e})$, we prove that
cycles can be characterised in the following way.
\begin{lemma}
  A set $T\subseteq\delta$ is not a cycle if and only if there exists a set of
  states $S\subseteq Q$ such that 
  \begin{enumerate}
  \item some transition $t\in T$ leads from a state in $S$ to a state
    outside $S$ and
  \item there is no transition in $T$ leading from a state outside $S$
    to a state in $S$.
  \end{enumerate}
  \todo{Nebo neprazdna cast $T$ vede z $S$ do $S$ a jina neprazdna
    cast $T$ vede z doplňu $S$ do doplňku $S$ a žádná hrana to
    nekrosí.}
\end{lemma}

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[smallautomaton]
    \begin{scope}[initial angle=90]
      \node[state,initial] (p) at (0,0) {};
      \node[state] (q) at (2,0) {};
      \path[->,auto]
      (p) edge[out=-150,in=150,looseness=10] pic {acc={0}{0}} pic {l=$1$} (p)
      (p) edge[out=30,in=150] pic {acc={1}{1}} pic {l=$2$} (q)
      (q) edge[out=30,in=-30,looseness=10] pic[pos=.33] {acc={2}{2}} pic[pos=.67] {acc={3}{3}} pic {l=$3$} (q)
      (q) edge[out=-150,in=-30] pic {acc={3}{3}} pic {l=$4$} (p);

      \node[acclabel] at (1,-1.5)
      {$(\Fin\tacc{1}{1}\wedge\Inf\tacc{2}{2})\vee(\Inf\tacc{0}{0}\wedge\Fin\tacc{1}{1}\wedge\Inf\tacc{3}{3})$};
    \end{scope}

% \begin{scope}[initial angle=90]
% \node[state,initial] (phi) at (0,0) {$\varphi\vphantom{\G}$};
% \node[state] (Fb) at (1.5,-1.5) {$\F b$};
% \path[->,auto]
% (phi) edge[loop left]
%   pic {acc={3}{1}} pic {l=$\true$} (phi)
% (phi) edge[loop right]
%   pic {acc={1}{3}} pic {l=$a$} (phi)
% (phi) edge[out=-135,in=-105,looseness=10]
%   pic[pos=.12] {acc={2}{2}} pic[pos=.3,left] {l=$b$} (phi)
% %(phi) edge[out=-45,in=45] (Fb)
% (phi) edge[out=-45,in=105]
%   node[above right,inner sep=.3,pos=.45] {$\bar{b}$} (Fb)
% (phi) edge[out=-45,in=-75,looseness=10]
%   pic[pos=.12] {acc={2}{2}} (phi)
% (Fb) edge[loop right]
%   pic {acc={3}{1}} pic {l=$\bar{b}$} (phi)
% (Fb) edge node {$b$} + (0,-1.1)
% ;
% \node[acclabel] at (.75,-3.2)
%   {$\Fin\tacc{3}{1}\wedge(\Fin\tacc{2}{2}\vee\Fin\tacc{1}{3})$};
% \end{scope}

% \begin{scope}[xshift=4.8cm,initial angle=90]
% \node[state,initial] (phi) at (0,-.7) {$\varphi\vphantom{\G}$};
% \path[->,auto]
% (phi) edge[loop left] 
%   pic {acc={3}{1}} pic {l=$\true$} (phi)
% (phi) edge[loop right]
%   pic {acc={1}{3}} pic {l=$a$} (phi)
% (phi) edge[out=-135,in=-105,looseness=10]
%   pic[pos=.75] {accsq={0}{5}} pic[pos=.15] {acc={2}{2}}
%   pic[left] {l=$b$} (phi)
% (phi) edge[out=-45,in=-75,looseness=10]
%   pic[pos=.15] {acc={2}{2}} pic[pos=.75] {acc={4}{4}}
%   pic[right] {l=$\bar{b}$} (phi)
% ;
% \node[acclabel,align=center] at (0,-2.8)
%   {$\Fin\tacc{3}{1}\wedge{}$\\$(\Fin\tacc{2}{2}\vee\Fin\tacc{1}{3})
%    \wedge{}$\\$(\Fin\tacc{4}{4}\vee\Inf\taccsq{0}{5})$};
% \end{scope}

  \end{tikzpicture}
  \caption{An automaton showing different reduction result for various reduction levels. \todo{Overit a dodelat!}}
  \label{fig:levels}
\end{figure}
